---
title: "AlberdiLab | EHI Squamate Rodent metagenomics"
subtitle: "EHI Squamate Rodent"
author:
  - Raphael Eisenhofer^[University of Copenhagen, raphael.eisenhofer@sund.ku.dk]

  - Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/EHI_squamate_rodent
description: |
  Data analysis code for the study on comparing squamates and rodent gut metagenomes
link-citations: yes
github-repo: alberdilab/EHI_squamate_rodent
---

# Introduction


```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```


This webbook contains all the code used for data analysis in study on comparing squamate and rodent gut metagenomes from the Earth Hologenome Initiative.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/EHI_squamate_rodent.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(rmarkdown)
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(broom)
library(broom.mixed)
library(janitor)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
library(plotly)
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtreeExtra)
library(ggtree)
library(ggh4x)

# For statistics
library(spaa)
library(vegan)
library(Rtsne)
library(geiger)
library(hilldiv2)
library(distillR)
library(ANCOMBC)
library(lme4)
library(matrixStats)
```
