---
title: "0_host_species_representation"
author: "Raphael Eisenhofer"
format: 
  gfm:
    fig-width: 12
    fig-height: 7
    fig-dpi: 300
    toc: true
  html:
    fig-width: 12
    fig-height: 7
    fig-dpi: 300
    toc: true
editor: visual
---

## Representativeness of the different host metagenomes

```{r}
#| message: false
#| echo: false

library(tidyverse)
library(ggrepel)
library(rairtable)
library(ggh4x)

coassembly_df <- read_delim("../data/EHI_squamate_rodent_coassembly_info.txt")

n_hosts <- nrow(coassembly_df)
n_squamata <- coassembly_df %>% filter(host_order == "squamata") %>% nrow(.)
n_rodent <- n_hosts - n_squamata
```

We have `r n_hosts` different host taxa, `r n_squamata` are squamates, `r n_rodent` are rodents.

How well represented are the host orders by metagenomic data and number of MAGs?

```{r}
#| message: false

set.seed(1337)

coassembly_df %>%
  ggplot(aes(x = host_order, y = n_mags, 
             colour = host_order, label = paste(host, n_samples))) +
  geom_jitter(height = 0, width = 0.1, alpha = 0.6, 
              aes(size = metagenomic_data_gbp)) + 
  geom_text_repel() + 
  theme_classic() +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```

## DAMR estimates by host order and species

Determining how representative the MAGs are of each host species using the Domain-Adjusted Mapping Rate (DAMR).

```{r}
#| message: false
#| warning: false


#Select the right view id
view <- airtable(base = 'appWbHBNLE6iAsMRV',
                 table = 'tblWDyQmM9rQ9wq57',
                 view = 'viwQ4Ow667d1lqFpf')

#wrangle
airtable_data_dmb <- read_airtable(view, 
                               id_to_col = TRUE, 
                               max_rows = 50000) %>%
  filter(DM_batch_static == "DMB0173") %>%
  select(EHI_sample_static, MAG_mapping_percentage,
         SingleM_fraction, host_species) %>%
  mutate(SingleM_fraction = as.numeric(SingleM_fraction),
         host_species = as.character(host_species),
         DAMR = MAG_mapping_percentage / SingleM_fraction) %>%
  inner_join(., coassembly_df, by = join_by(host_species == host))

#plot
airtable_data_dmb %>%
  filter(EHI_sample_static != "EHI01045" & EHI_sample_static != "EHI01812") %>%
  ggplot(aes(x = host_species, y = DAMR, 
             colour = host_order, label = host_species)) +
  geom_jitter(
    size = 3, 
    height = 0, 
    width = 0.15, 
    alpha = 0.5
  ) +
  facet_nested(~host_order, scale = "free") +
  stat_summary(
    fun = "mean", 
    geom = "crossbar", 
    colour = "red", 
    width = 0.5
  ) +
  stat_summary(
    fun = "mean",
    geom = "text",
    aes(label = round(..y.., 2)),
    vjust = -1,
    colour = "black"
  ) +
  scale_y_continuous(limits = c(0.3, 1.3)) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    axis.text.x = element_text(angle = 90, face = "italic",
                               hjust = 1, vjust = 0.25),
    legend.position = "none"
    )
  

```

Note I've filtered out both EHI01045 (no SMF value) and EHI01812 (only 0.16% mapping).
